### 2020-01-16 ISSTA/concepts.tex
big pass

We define two basic types of determinism, shown in Figure
\ref{fig:types}:  horizontal determinism and vertical determinism.  In
horizontal determinism, which is what we usually think of when we
think about deterministic behavior, a software system reliably
produces the same behavior given the same steps of a test: the
behavior of multiple executions that are ``the same'' in terms of
inputs/actions can be aligned and checked for some type of equality.  In
vertical determinism, rather than behavior across executions, we are
interested in behavior within an execution, where repeating the same
step of a test twice should result in the same behavior.  Obviously,
vertical determinism is both rarer and more specialized than
horizontal determinism.  However, in those instances, vertical
determinism is arguably more important than horizontal determinism, in
that violations usually indicate a potentially serious fault.

@@

Horizontal determinism can be best understood by thinking of
nondeterminism detection as an unusual kind of \emph{differential
  testing} \cite{Differential,ICSEDiff}.  In differential testing, a
system is compared against a reference in order to ensure that
it behaves equivalently, at some level of abstraction, to another
implementation of the same functionality.  Differential testing is
extremely powerful, in that any divergence of behavior, if divergence
is correctly defined, indicates a functional correctness fault in (at
least) one of the systems under test, and  is widely
used for systems software components such as compilers
\cite{Differential,csmith} and POSIX file systems \cite{CFV08,AMAI}, where multiple
implementations are common.  The major limitation of differential
testing is that multiple implementations of a system are
almost as rare as good correctness specifications.

For the special case of detecting nondeterminism, however, \emph{a
  system can serve as its own reference implementation.}    The
problem, then, becomes one of deciding at what granularity the
reference equivalence will be checked:  e.g., 
processor-instruction and memory-layout determinism is seldom
necessary or even desired.  We propose two approaches to ``aligning'' an
execution with itself.

@@

We can formally distinguish the types of nondeterminism by using a
variant of a labeled transition system (LTS) as our model, where (1) $S$
is a set of states; (2) $V$ is a set of \emph{observable} states, where
$|V| \leq |S|$; (3) $v: S \rightarrow V$ is a total function that, given a state, maps it
  into the set of observable states, such that every state has an observable
  component, which may be the complete state, or only an aspect of the
  full state; (4)  $I \subseteq S$ is a set of initial states; (5) $A$
  is a set of \emph{actions}; and (6) $T \subseteq S \times A \times S$ is a transition relation.

@@

We assume that the underlying behavior of a system may be
deterministic, or nondeterministic, in the strict sense, by allowing
for a transition relation that \emph{may} be a function of
$S \times A$.  A \emph{trace} $t$ is a finite sequence $t = s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall i < n
. (s_i, \alpha_i, s_{i+1}) \in T$.  The concept of an action does not restrict this approach to reactive
systems; it can be generalized to consider the only action that varies
to be the selection of an input, $\alpha_0$, with the remainder of the
actions being internal $\tau$ actions, in the usual fashion.

@@

A pair of traces $(t_1, t_2)$ where $t_1 = s^1_0
\xrightarrow[]{\alpha^1_0} s^1_1 \xrightarrow[]{\alpha^1_1} \ldots
\xrightarrow[]{\alpha^1_{n-1}} s^1_n$ and $t_2 = s^2_0
\xrightarrow[]{\alpha^2_0} s_1 \xrightarrow[]{\alpha^2_1} \ldots
\xrightarrow[]{\alpha^2_{n-1}} s^2_n$ are said to show \emph{visible value
  nondeterminism} if $\exists i > 0$ such that: $\forall j < i . \alpha^1_j = \alpha^2_j \wedge
v(s^1_j) = v(s^2_j)$ but $v(s^1_i) \neq v(s^2_i)$.  \emph{Final state
  nondeterminism} is defined in the same way, except with the
restriction that $i = n$.  A pair of traces may exhibit visible value
nondeterminism but not final state nondeterminism, but any pair of
traces exhibiting final state nondeterminism also demonstrate visible
value nondeterminism.

@@

To define vertical nondeterminism with
respect to idempotency, we can
define $\mathcal{I} \subseteq A$, the set of \emph{idempotent}
actions.  A trace $t s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ shows \emph{vertical nondeterminism
  with respect to idempotent operations} if $\exists i < n$ such that
$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
\neq v(s_{i+2})$.  Note that vertical nondeterminism may, in theory, exhibit only
after a sequence of more than two successive applications of a (supposedly)
idempotent operation; the definition only requires that after \emph{some}
sequence of actions, possibly with $\alpha_{i-1} = \alpha_i$, the
visible state changes.  In an implementation, we would usually restrict the
search to a finite number of repetitions; for many faults, two
``copies'' will suffice, since an operation is either truly idempotent or
not idempotent.
%Only the existence of \emph{non-visible} state changes that
%can eventually produce visible state changes \emph{after a succession of the
%same action} require more repetitions.

@@

In order to implement a practical nondeterminism-detection tool, it is
important to consider the real-world sources of nondeterminism.  
\begin{comment}
Most of these
sources can, in principle, be detected simply by re-running a test
twice and comparing results, either at all visible values or at the
final state.  For example, if some library call uses a random number
generator (perhaps in a stochastic algorithm), the returned value should be
different on a second run.  Timing dependencies may change on a second
execution, especially if a delay is added between steps of the test (a
more principled version of the unfortunate tradition of testing
concurrent code by adding ad-hoc {\tt sleep} statements).
\end{comment}
Most nondeterministic behavior in tests is probably attributable to simple
effects that can vary with a repeated execution, in the same
(approximate) enviroment, of the same test.  The two broad classes of
nondeterminism that can usually be detected by
executing the same test twice in succession, without attempting
to control any other factors, are:

\begin{itemize}
\item {\bf External Environment:} This class of nondeterminism arises
  when a test's behavior depends on factors outside the program under
  test, which are subject to uncontrolled variance.  Calling {\tt
    random()} in Python is an obvious example. Similarly, calls to {\tt time}
  functions  will usually return different values on different runs,
  and even if only elapsed time is used, due to system load, I/O, and
  other external factors.
  %Finally, in
  %distributed system testing, the environment is often remote servers
  %(or clients) that may be unavailable, slower than expected, or
  %subject to faulty behavior for reasons not under the control of the
  %test.
\item {\bf Concurrency:} A second common cause for
  nondeterminism in tests is when an algorithm is implemented using
  concurrency, in the (incorrect) belief that the result of the
  algorithm remains deterministic.  E.g., a multi-threaded
  implementation of a search algorithm, that returns the location of
  an item in an unsorted list; in the presence of duplicate items, thread scheduling may change the index returned.
  %If the list may contain duplicates,
  %and a developer has not accounted for this possibility, the
  %algorithm may return different locations of an item
  %depending on the scheduling of threads.
  %Arguably, this can be
  %subsumed under the first class, with the behavior of the
  %processor/scheduler as the ``enviromental'' factor; relying on
  %scheduling is, in a sense, equivalent to calling {\tt random} or
  %{\tt time}.
\end{itemize}

Lam et al. \cite{idflakies} note that flaky tests not due to order
dependence are generally due to ``concurrency, timeouts,
network/IO, etc.'' a description that is well accounted for by these two
primary sources, and most of which can be detected easily
%, at least in principle, 
by multiple runs of a test.
%However, simply running a test twice in the same process (possibly with a delay between steps) does
%not suffice in all cases.

@@

Delta-debugging \cite{DD}\footnote{We use delta-debugging to stand for
  all test reduction algorithms, even those \cite{CReduce,tstl} that do not use the exact
  binary-search that distinguishes delta-debugging \emph{per se}; the
  differences are immaterial for our purposes.}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
The core idea of delta-debugging is to take a test that satisfies some
property (usually the property is ``it fails in a particular way'')
and produce a smaller test that satisfies the same property.
Delta-debugging, as proposed by Zeller and Hildebrandt, relies on a
modified binary search, but the
following discussion relies only on the general structure of
delta-debugging \cite{CReduce,onetest}, which is as follows.  Given a
test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
$\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$, we
reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$
as follows:

\begin{enumerate}
\item $T_{\mathit{CURR}} = T_{\mathit{INIT}}$
\item Let $T_{\mathit{NEXT}} =$ a variation of $T_{\mathit{CURR}}$
  that has not yet been tried.  If there are no further variations of
  $T_{\mathit{CURR}}$, stop and return $T_{\mathit{CURR}}$ as the
  reduced test.
\item Add $T_{\mathit{NEXT}}$ to the set of variations that have been
  tried.
\item If $\mathit{PRED}(T_{\mathit{NEXT}})$, set $T_{\mathit{CURR}} =
  T_{\mathit{NEXT}}$.
\item Go to 2.
\end{enumerate}

@@

Delta-debugging in the context of detecting nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls (it is one of these ten
calls), but very difficult in a test consisting of a hundred library
function calls.  This is the typical use of
delta-debugging.  However, in horizontal nondeterminism detection,
delta-debugging also tends to \emph{change the probability of
  nondeterministic behavior}; this can be both harmful and
beneficial.  The behavior is part of a more general (and, to our
knowledge, not previously investigated) issue:  using delta-debugging
to minimize a test with respect to a predicate $\mathit{PRED}$ that only holds with a
certain probability (the predicate itself is nondeterministic).  Note
that while most descriptions of delta-debugging discuss reducing a
test with respect to \emph{test failure} (hence the term
``delta-\emph{debugging}'') we adopt the wider view of delta-debugging
as reducing the length of a test while maintaining that an arbitrary
predicate still holds.   E.g., predicates, related to code-coverage can be very useful
\cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution is to consider
what happens when $\mathit{PRED}$ is not deterministic.
``The test behaves flakily'' is a simple, important, example of such a $\mathit{PRED}$.

In \emph{monotonic} probabilistic delta-debugging, removing part of a test \emph{cannot
increase the probability of the predicate holding}.  This is fairly
common.  In these cases, when we use a reduction algorithm to reduce $t$
to $r$, $P(\mathit{PRED}(r)) \leq P(\mathit{PRED}(t))$.  In \emph{non-monotonic} cases,
however, there is no such upper bound.  Removing a step in $t$ may
\emph{increase} the probability that $t$ behaves nondeterministically.
\begin{comment}
The $\mathit{PRED}$ of interest for horizontal determinism is basically of
the form:  {\bf the test will exhibit at least two different behaviors in $S$
runs, with probability $P$}.  This predicate may be monotonic or
non-monotonic, depending on the causes of the nondeterminism detected.
\end{comment}
In probabilistic
settings, we may want to exploit non-monotonicity to find a
subset of a test where a predicate that is very far from holding (in terms of probability, $P$) on the original test now
holds.  For instance, a test may be very infrequently nondeterministic, and we want to find a subset where there is much greater probability of variance.

@@

\begin{figure}
\centering 
\begin{subfigure}{0.30\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddmin}
\caption{Standard delta-debugging}
\label{fig:p1}
\end{subfigure}
\begin{subfigure}{0.30\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddminforcep}
\caption{P=0.5, 100 samples}
\label{fig:p2}
\end{subfigure}
\begin{subfigure}{0.30\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddminforceprep}
\caption{P=0.5, 10 samples, 10 replications}
\label{fig:p3}
\end{subfigure}
\caption{Delta-debugging of the same simple test}
\end{figure}

However, simply using delta-debugging off the shelf with a $\mathit{PRED}$
such as $P(fail) > 0.3 \wedge P(fail) < 0.7$, to force a test to be highly flaky, and force it to fail
sufficiently often to be used in debugging, will often produce
surprising and unfortunate results.  In many cases,
delta-debugging will indeed reduce the large test to a small subset.
And, in a technical sense, delta-debugging will work:  it will never
convert a nondeterministic test to a completely deterministic test,
because the reduced test $r$ that delta-debugging returns is always one
such that $\mathit{PRED}(r)$ has evaluated to true at least once.
However, if you run the resulting test, it will, in many cases, have a
$P(fail)$ that is much, much smaller than
0.3, perhaps as low as 0.01.  Why?

The problem is analogous to the problem of publication bias in
scientific fields \cite{ahmed2012assessment}.  We can think of each
application of $\mathit{PRED}$ to a candidate variant
$T_{\mathit{NEW}}$ as a scientific experiment.  A predicate like
$P(fail) > 0.3 \wedge P(fail) < 1.0$ cannot be evaluated by
determining the true probability of failure; we have no more direct access to
this value than a medical researcher has access to the true effect
size and direction of a proposed medical treatment; rather, the test must be
run some concrete number of times, and the number of failures counted.
Even if the number of samples $N$ is large, there is some
probability (based on the sample size) of a result that diverges
significantly from the actual probability of failure.  If the
predicate were run only once, and the number of samples reasonably
large, this would not matter.  However, by their nature,
delta-debugging and other test reduction algorithms explore a search
space that often contains thousands or even millions of
candidate tests.  The predicate is evaluated on each of these, and so
even with large $N$, it is extremely likely that \emph{some} evaluation will
produce a very poor estimate of $P(fail)$.  At this point, delta-debugging is
``stuck'' with the error, because test reduction
algorithms usually do not allow backtracking.  After such a ``mistake,'' finding further
reductions will become harder, but may still be possible due to the
same source of errors:  \emph{the number of experiments run is far larger
  than the probability of an incorrect result for a single
  experiment}.  We define a \emph{false positive} in probabilistic
delta debugging as a case where $\mathit{PRED}$ evaluates to true, but
the actual probability constraints of $\mathit{PRED}$ are not satsified.  False negatives
also exist, but are usually harmless, since it often just forces a different, equally
desirable reduction path.

The analogy to publication bias is simple.  Assume that a paper in,
e.g., a medical journal will only be published if it shows that a
treatment is effective, with the statistical requirement that $p <
0.05$ (that is, by some statistical test, the probability
that the result is due to chance rather than reflecting reality is
less than 5\%).  Even if no researchers are dishonest, 5 of every 100 papers published
in the journal will be report an effect with the wrong effect size or
even effect direction!  Now consider a ``hot'' topic, a kind of
treatment many research teams may investigate.  Major journals tend
not to publish papers that say ``this new treatment, which is not
established, may not work,'' (especially if the result is not ``the
treatment certainly does not work'' but only ``the experiment did not
establish $p < 0.05$''), but they  are very likely to publish a paper
that says ``this exciting new treatment works!''  In addition to
sometimes producing excitement about treatments that do not work, this
bias tends to magnify the expected effect size of treatments, since
the chance of getting a ``passing'' $p$-value is higher when an
experiment finds a larger effect size, due to how most statistical
tests work.
The \emph{bias} in
favor of positive results is shared by delta-debugging;
delta-debugging's behavior is much more influenced by cases where
$\mathit{PRED}$ holds than cases where it does not.  Looking at the
literature, the unpublished papers are invisible, magnifying the
effect on scientific opinion of the published experiments.

@@

In scientific literature, the most frequently proposed solution is the
use of replications:  repeated runs of ``successful'' experiments to
minimize the probability that a direction or effect size is a fluke due to publication
bias.  One way to produce this effect would be to allow
delta-debugging to backtrack if the probabilities observed in predicate
evaluations suddenly exhibit a strong discontinuity, a kind of ``paper
retraction'' based on near-replications.  However, this requires
modifying delta-debugging implementations, which is difficult and
sometimes not really feasible (e.g., few users of afl-fuzz \cite{aflfuzz} will wish
to change its C code for test reduction).  Ideally, the solution
should be implementable simply by modifying $\mathit{PRED}$.

A costly but somewhat effective solution is to make $N$ large in comparison to
the number of expected predicate evaluations performed during
delta-debugging.  However, given the large number of evaluations performed,
this will tend to make delta-debugging extremely slow.  We propose
using a dynamic sampling approach, where $N$ is small, but if the
predicate evaluates to true, $M$ repeated true evaluations
(``replications'', where the first true evaluation is counted as 1 replication) are
required before the predicate returns ``true'' to the delta-debugging
algorithm.  To evaluate the predicate $\mathit{PRED}$ given $N$,
$M$, and desired probability threshold $p$, the algorithm is simple:

\begin{enumerate}
  \item  $i = 0$
  \item Evaluate $\mathit{PRED}$ $N$ times.  If the proportion of cases
  where $\mathit{PRED}$ was true is less than $p$, return false.
  \item  $i = i + 1$
  \item If $i < M$ go to step 1.
  \item  Return true.
\end{enumerate}

A set of $M$ repeated false positives with
$N = \frac{K}{M}$ samples each is much less likely than a false positive
with $K$ samples; so long as we accept the resulting bias in favor of
false negatives, we can therefore produce a reduced test with a
desired $P(fail)$ much more cheaply:  the use of replications not only
means we only pay the full sampling price on rare occasions, but a
desired accuracy for $P$ can be obtained with a much smaller value $M
\times N$ than a non-dynamically-sampled $N$.  For example, if $\mathit{PRED}$
is $P(fail) \geq 0.5$, and the true probability for a candidate test
is 0.25, using $N=8$ will give a false positive rate of over 10\%.
Using 4 replications on just 2 samples ($M=4;N=2$) yields a false
positive rate of about 3.7\%, yet requires almost 60\% fewer
executions of the test.  The probability calculations for such comparisons are
relatively simple, but in real testing are
usually not very useful, since the true
probability distributions of test behaviors vary widely and
dynamically during the
delta-debugging process; gathering experimental data during a trial
delta-debugging run and tuning $N$ and $M$ to yield desired
results is  more effective.  In fact, by tuning $M$, any degree of confidence can be achieved, with
the basic tradeoff being between finding a test with the desired
probabilities, and the speed and effectiveness of delta-debugging.
Because increasing $M$ makes false negatives more likely, larger $M$
will usually result in less-than-optimal reduction of the original
test.

>>>>

We define two basic types of determinism, shown in Figure
\ref{fig:types}:  horizontal determinism and vertical determinism.  In
horizontal determinism a software system reliably
produces the same behavior given the same steps of a test: the
behavior of multiple executions that are ``the same'' in terms of
inputs/actions can be aligned and checked for equality.  In
vertical determinism, rather than behavior across executions, we are
interested in behavior within an execution, where repeating the same
step of a test twice should result in the same behavior.  Vertical determinism is both rarer and more specialized than
horizontal determinism.  However, vertical
determinism violations usually indicate a potentially serious fault.

@@

Horizontal determinism can be best understood by thinking of
nondeterminism detection as an unusual kind of \emph{differential
  testing} \cite{Differential,ICSEDiff}.  In differential testing, a
system is compared against a reference in order to ensure that
it behaves equivalently, at some level of abstraction, to another
implementation.  Differential testing is
extremely powerful, in that any (properly defined) divergence of
behavior indicates a functional correctness fault in (at
least) one of the systems under test, and  is widely
used for systems software such as compilers
\cite{Differential,csmith} and file systems \cite{CFV08,AMAI}.  The major limitation of differential
testing is that multiple implementations of a system are
almost as rare as good correctness specifications.
For the special case of detecting nondeterminism, however, \emph{a
  system can serve as its own reference implementation.}    The
problem, then, becomes one of deciding at what granularity the
reference equivalence will be checked:  e.g., 
processor-instruction and memory-layout determinism is seldom
necessary or even desired.  We propose two approaches to ``aligning'' an
execution with itself.

@@

We can formally distinguish the types of nondeterminism by using a
variant of a labeled transition system (LTS)l, where (1) $S$
is a set of states; (2) $V$ is a set of \emph{observable} states, where
$|V| \leq |S|$; (3) $v: S \rightarrow V$ is a total function that, given a state, maps it
  into the set of observable states, such that every state has an observable
  component, which may be the complete state, or only an aspect of the
  full state; (4)  $I \subseteq S$ is a set of initial states; (5) $A$
  is a set of \emph{actions}; and (6) $T \subseteq S \times A \times S$ is a transition relation.

@@

We assume that the \emph{underlying} behavior of a system may be
deterministic, or nondeterministic  by allowing
for a transition relation that \emph{may} be a function of
$S \times A$.  A \emph{trace} $t$ is a finite sequence $t = s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall i < n
. (s_i, \alpha_i, s_{i+1}) \in T$.  The concept of an action does not restrict this approach to reactive
systems; it can be generalized to consider the only action that varies
to be the selection of an input, $\alpha_0$, with the remainder of the
actions being internal $\tau$ actions.

@@

A pair of traces $(t_1, t_2)$ where $t_1 = s^1_0
\xrightarrow[]{\alpha^1_0} s^1_1 \xrightarrow[]{\alpha^1_1} \ldots
\xrightarrow[]{\alpha^1_{n-1}} s^1_n$ and $t_2 = s^2_0
\xrightarrow[]{\alpha^2_0} s_1 \xrightarrow[]{\alpha^2_1} \ldots
\xrightarrow[]{\alpha^2_{n-1}} s^2_n$ are said to show \emph{visible value
  nondeterminism} if $\exists i > 0$ such that: $\forall j < i . \alpha^1_j = \alpha^2_j \wedge
v(s^1_j) = v(s^2_j)$ but $v(s^1_i) \neq v(s^2_i)$.  \emph{Final state
  nondeterminism} is defined in the same way, except with the
restriction that $i = n$.  A pair of traces may exhibit visible value
nondeterminism but not final state nondeterminism.

@@

To define vertical nondeterminism with
respect to idempotency, we can
define $\mathcal{I} \subseteq A$, the set of \emph{idempotent}
actions.  A trace $t s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ shows \emph{vertical nondeterminism
  with respect to idempotent operations} if $\exists i < n$ such that
$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
\neq v(s_{i+2})$.  Note that vertical nondeterminism may, in theory, exhibit only
after a sequence of more than two successive applications of a (supposedly)
idempotent operation; the definition only requires that after \emph{some}
sequence of actions, possibly with $\alpha_{i-1} = \alpha_i$, the
visible state changes.  In an implementation, we would usually restrict the
search to a finite number of repetitions; for many faults, two
``copies'' will suffice.
%Only the existence of \emph{non-visible} state changes that
%can eventually produce visible state changes \emph{after a succession of the
%same action} require more repetitions.

@@

In order to implement a practical nondeterminism-detection tool, it is
important to consider the real-world sources of nondeterminism.  
\begin{comment}
Most of these
sources can, in principle, be detected simply by re-running a test
twice and comparing results, either at all visible values or at the
final state.  For example, if some library call uses a random number
generator (perhaps in a stochastic algorithm), the returned value should be
different on a second run.  Timing dependencies may change on a second
execution, especially if a delay is added between steps of the test (a
more principled version of the unfortunate tradition of testing
concurrent code by adding ad-hoc {\tt sleep} statements).
\end{comment}
Most nondeterministic behavior in tests is probably attributable to simple
effects that can vary with a repeated execution, in the same
(approximate) enviroment, of the same test.  There are two broad classes of
nondeterminism that can usually be detected by
executing the same test twice in succession, without attempting
to control any other factors.  {\bf External Environment} nondeterminism arises
  when a test's behavior depends on factors outside the program under
  test, which are subject to uncontrolled variance.  Calling {\tt
    random()} in Python is an obvious example. Similarly, calls to {\tt time}
  functions  will usually return different values on different runs,
  and even if only elapsed time is used, due to system load, I/O, and
  other external factors.  {\bf Concurrency} is a second common cause for
  nondeterminism, when an algorithm is implemented using
  concurrency, in the (incorrect) belief that the result of the
  algorithm remains deterministic.  E.g., a multi-threaded
  implementation of a search algorithm, that returns the location of
  an item in an unsorted list; in the presence of duplicate items, thread scheduling may change the index returned.

\begin{comment}
\begin{itemize}
\item {\bf External Environment:} This class of nondeterminism arises
  when a test's behavior depends on factors outside the program under
  test, which are subject to uncontrolled variance.  Calling {\tt
    random()} in Python is an obvious example. Similarly, calls to {\tt time}
  functions  will usually return different values on different runs,
  and even if only elapsed time is used, due to system load, I/O, and
  other external factors.
  %Finally, in
  %distributed system testing, the environment is often remote servers
  %(or clients) that may be unavailable, slower than expected, or
  %subject to faulty behavior for reasons not under the control of the
  %test.
\item {\bf Concurrency:} A second common cause for
  nondeterminism in tests is when an algorithm is implemented using
  concurrency, in the (incorrect) belief that the result of the
  algorithm remains deterministic.  E.g., a multi-threaded
  implementation of a search algorithm, that returns the location of
  an item in an unsorted list; in the presence of duplicate items, thread scheduling may change the index returned.
  %If the list may contain duplicates,
  %and a developer has not accounted for this possibility, the
  %algorithm may return different locations of an item
  %depending on the scheduling of threads.
  %Arguably, this can be
  %subsumed under the first class, with the behavior of the
  %processor/scheduler as the ``enviromental'' factor; relying on
  %scheduling is, in a sense, equivalent to calling {\tt random} or
  %{\tt time}.
\end{itemize}
\end{comment}

Lam et al. \cite{idflakies} note that flaky tests not due to order
dependence are generally due to ``concurrency, timeouts,
network/IO, etc.'' a description that is well accounted for by these two
primary sources.
%, and most of which can be detected easily
%, at least in principle, 
%by multiple runs of a test.
%However, simply running a test twice in the same process (possibly with a delay between steps) does
%not suffice in all cases.

@@

Delta-debugging \cite{DD}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
The core idea of delta-debugging is to take a test with some
property (usually ``it fails'')
and produce a smaller test that has the same property.
Delta-debugging as originally proposed uses a
modified binary search, but we rely only on the general structure: \cite{CReduce,onetest}:  Given a
test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
$\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$, we
reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$
as follows:

\begin{enumerate}
\item $T_{\mathit{CURR}} = T_{\mathit{INIT}}$
\item Let $T_{\mathit{NEXT}} =$ a variation of $T_{\mathit{CURR}}$
  that has not yet been tried.  If none exist, stop and return $T_{\mathit{CURR}}$.
\item Add $T_{\mathit{NEXT}}$ to the set of variations that have been
  tried.
\item If $\mathit{PRED}(T_{\mathit{NEXT}})$, set $T_{\mathit{CURR}} =
  T_{\mathit{NEXT}}$.
\item Go to 2.
\end{enumerate}

@@

Delta-debugging in the context of nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls but very difficult with
more than a hundred calls.  In horizontal nondeterminism detection, however,
delta-debugging also tends to \emph{change the probability of
  nondeterministic behavior}; this can be both harmful and
beneficial.  This behavior is part of a more general (and, to our
knowledge, not previously investigated) issue: reducing a test with
respect to an arbitrary predicate $\mathit{PRED}$ (including failure,
but also code coverage, etc. \cite{icst2014,stvrcausereduce,NonAdeq}) that \emph{only holds with a
certain probability} (the predicate itself is nondeterministic).
\begin{comment}Note
that while most descriptions of delta-debugging discuss reducing a
test with respect to \emph{test failure} (hence the term
``delta-\emph{debugging}'') we adopt the wider view of delta-debugging
as reducing the length of a test while maintaining that an arbitrary
predicate still holds.   E.g., predicates, related to code-coverage can be very useful
\cite{icst2014,stvrcausereduce,NonAdeq}.
\end{comment}
%We want to consider
%what happens when $\mathit{PRED}$ is not deterministic.
``The test behaves flakily'' is an obvious relevant example of such a $\mathit{PRED}$.

For \emph{monotonic} $\mathit{PRED}$, removing part of a test \emph{cannot
increase the probability of the predicate holding}: when we reduce $t$
to $r$, $P(\mathit{PRED}(r)) \leq P(\mathit{PRED}(t))$.  This is fairly
common.  In \emph{non-monotonic} cases,
however, there is no such upper bound.  Removing a step in $t$ may
\emph{increase} the probability of $\mathit{PRED}$.
\begin{comment}
The $\mathit{PRED}$ of interest for horizontal determinism is basically of
the form:  {\bf the test will exhibit at least two different behaviors in $S$
runs, with probability $P$}.  This predicate may be monotonic or
non-monotonic, depending on the causes of the nondeterminism detected.
\end{comment}
In probabilistic
settings, we can often exploit non-monotonicity.  For instance, a test may be
``almost non-flaky'' and we want $P(\mathit{fail})$ to be closer to
50\%, to aid debugging.

@@

\begin{figure}
\centering 
\begin{subfigure}{0.30\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddmin}
\caption{No modification}
\label{fig:p1}
\end{subfigure}
\begin{subfigure}{0.30\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddminforcep}
\caption{N=100}
\label{fig:p2}
\end{subfigure}
\begin{subfigure}{0.30\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddminforceprep}
\caption{M=10, N=10}
\label{fig:p3}
\end{subfigure}
\caption{Reducing the same simple test; x-axis is time/changing
  $T_\mathit{CURR}$; y-axis is true $P(\mathit{PRED})$.}
\end{figure}

However, simply using delta-debugging off the shelf with a $\mathit{PRED}$
such as $P(\mathit{fail}) > 0.3 \wedge P(\mathit{fail}) < 0.7$, to force a test to be highly flaky, and force it to fail
sufficiently often to be used in debugging, will often produce
surprising and unfortunate results.  In many cases,
delta-debugging will indeed reduce the large test to a small subset.
And, in a technical sense, delta-debugging will work:  it will never
convert a nondeterministic test to a completely deterministic test,
because the reduced test $r$ that delta-debugging returns is always one
such that $\mathit{PRED}(r)$ has evaluated to true at least once.
However, if you run the resulting test, it will, in many cases, have a
$P(\mathit{fail})$ that is much, much smaller than
0.3, perhaps as low as 0.01.  Why?

The problem is analogous to the problem of publication bias in
scientific fields \cite{ahmed2012assessment}.  We can think of each
application of $\mathit{PRED}$ to a candidate variant
$T_{\mathit{NEW}}$ as a scientific experiment.  A predicate like
$P(\mathit{fail}) > 0.3 \wedge P(\mathit{fail}) < 1.0$ cannot be evaluated by
determining the true probability of failure; we have no more direct access to
this value than a medical researcher has access to the true effect
size and direction of a proposed medical treatment; rather, the test must be
run some concrete number of times, and the number of failures counted.
Even if the number of samples $N$ is large, there is some
probability (based on the sample size) of a result that diverges
significantly from the actual probability of failure.  If the
predicate were run only once, and the number of samples reasonably
large, this would not matter.  However, test reduction algorithms explore a search
space that often contains thousands or millions of
tests.  The predicate is evaluated on each of these, and so
even with large $N$, it is likely that \emph{some} evaluation will
produce a poor estimate of $P(\mathit{fail})$.  At this point, reduction is
``stuck'' with the error, because the
algorithms usually do not allow backtracking.  After such a ``mistake,'' finding further
reductions will become harder, but may still be possible due to
another unlucky evaluation.  We define a \emph{false positive} in probabilistic
delta debugging as a case where $\mathit{PRED}$ evaluates to true, but
the actual probability constraints of $\mathit{PRED}$ are not
satsified (false negatives
also exist, but are less of a problem).

The analogy to publication bias is simple.  Assume that a paper in,
e.g., a medical journal will only be published if it shows that a
treatment is effective, with the statistical requirement that $p <
0.05$.  Even if no researchers are dishonest, 5 of every 100 papers published
in the journal will be report an effect with the wrong effect size or
even direction!  Now consider a ``hot'' topic, a kind of
treatment many research teams may investigate.  Major journals tend
not to publish papers that say ``this new treatment, which is not
established, may not work,'' (especially if the result is not ``we
accept the null hypothesis'' but only ``the experiment did not
show $p < 0.05$''), but they  are very likely to publish a paper
that says ``this exciting new treatment works!'' 
The \emph{bias} in
favor of positive results is shared by test reduction, which;
is much more influenced by cases where
$\mathit{PRED}$ holds than cases where it does not.  The ``unpublished papers'' are invisible, magnifying the
effect of the ``published'' experiments.

@@

In scientific literature, the most frequently proposed solution is the
use of replications:  repeated runs of ``successful'' experiments to
minimize the probability that a direction or effect size is a fluke.  One way to produce this effect would be to allow
reducers to backtrack if the probabilities observed in predicate
evaluations suddenly exhibit a strong discontinuity.  However, this requires
modifying the implementations, which is difficult and
sometimes not really feasible.  Ideally, the solution
should be implementable simply by modifying $\mathit{PRED}$ itself.

A costly but plausible solution is to make $N$ large in comparison to
the number of expected predicate evaluations performed during
delta-debugging.  However, given the large number of evaluations performed,
this will tend to make reduction extremely slow, since $N$ must be
very large indeed.  We propose
using a dynamic sampling approach, where $N$ is small, but if the
predicate evaluates to true, $M$ repeated true evaluations
(``replications'') are
required before the predicate is counted as holding.  To evaluate the predicate $\mathit{PRED}$ given $N$,
$M$, and desired probability bound $p$, the algorithm is:

\begin{code}
for $i = 0 \ldots M-1$
   $T =$ \# times $\mathit{PRED}$ is TRUE over $N$ evaluations.
   if $\frac{T}{N} < p$, return FALSE.
return TRUE
\end{code}

A set of $M$ repeated false positives with
$N = \frac{K}{M}$ samples each is much less likely than a false positive
with $K$ samples; so long as we accept the resulting bias in favor of
false negatives, we can therefore produce a reduced test with a
desired $P(\mathit{fail})$ much more cheaply:  the use of replications not only
means we only pay the full sampling price on rare occasions, but a
desired accuracy for $P$ can be obtained with a much smaller value $M
\times N$ than a non-dynamically-sampled $N$.  For example, if $\mathit{PRED}$
is $P(\mathit{fail}) \geq 0.5$, and the true probability for a candidate test
is 0.25, using $N=8$ will give a false positive rate of over 10\%.
Using 4 replications on just 2 samples ($M=4;N=2$) yields a false
positive rate of about 3.7\%, yet requires almost 60\% fewer
executions of the test.  The probability calculations for such comparisons are
relatively simple, but in real testing are
usually not very useful, since the true
probability distributions of test behaviors vary widely and
dynamically during the
delta-debugging process; gathering experimental data during a trial
delta-debugging run and tuning $N$ and $M$ to yield desired
results is  more effective.  In fact, by tuning $M$, any degree of confidence can be achieved, with
the basic tradeoff being between finding a test with the desired
probabilities, and the speed and effectiveness of reduction.
Because increasing $M$ makes false negatives more likely, larger $M$
will usually result in less-than-optimal reduction of the original
test.
